---
title: "Reproducible Research - Peer Assessment 2"
author: "chowchowsheep"
date: "December 5, 2018"
output: 
  html_document:
    keep_md: yes
---
## Downloading tools
```{r}
library(tidyverse)
```

## Downloading the file from source
```{r}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

download.file(fileURL, destfile = "storm.bz2")
storm <- read.csv("storm.bz2")
```

## Cleaning data

```{r}
storm.clean <- subset(storm, FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0) %>% as_tibble()
```

## Total fatalities and injuries by event type
```{r}
fatalities <- storm.clean %>% group_by(EVTYPE) %>%  summarize(Fatalities=sum(FATALITIES))

names(fatalities) <- c("Event", "Fatalities")
fatalities <- fatalities[order(fatalities$Fatalities,decreasing=TRUE),]

injuries <- storm.clean %>% group_by(EVTYPE) %>%  summarize(Injuries=sum(INJURIES))

names(injuries) <- c("Event", "Injuries")
injuries <- injuries[order(injuries$Injuries,decreasing=TRUE),]

fat.inj <- left_join(injuries,fatalities,by=c("Event"= "Event"))
fat.inj$Combined <- fat.inj$Injuries*0.5 + fat.inj$Fatalities
fat.inj <- fat.inj[order(fat.inj$Combined,decreasing=TRUE),]

```

## Cleaning damage data
### To account for the scaling factor, we multiplied the damage by the corresponding scale. For other factors which are non-standard, the recording might have been done incorrectly, so we use a factor of 1 for those observations
```{r}
storm1 <- storm.clean %>%
  mutate(PROPDMG = case_when(
    PROPDMGEXP == "K" ~ PROPDMG * 1e3,
    PROPDMGEXP == "M" ~ PROPDMG * 1e6,
    PROPDMGEXP == "B" ~ PROPDMG * 1e9,
    PROPDMGEXP == "" ~ PROPDMG * 1,
    TRUE ~ PROPDMG
  )) 
  
storm2 <- storm.clean %>%
  mutate(CROPDMG = case_when(
    CROPDMGEXP == "K" ~ CROPDMG * 1e3,
    CROPDMGEXP == "k" ~ CROPDMG * 1e3,
    CROPDMGEXP == "M" ~ CROPDMG * 1e6,
    CROPDMGEXP == "m" ~ CROPDMG * 1e6,
    CROPDMGEXP == "B" ~ CROPDMG * 1e9,
    CROPDMGEXP == "" ~ CROPDMG * 1,
    TRUE ~ CROPDMG
  ))
```

## Total property and crop damage by event type
### 1. Summing up the property and crop damage, based on event type, then joining the two dataframes. The total damage is calculated by summing the crop and property damage.
```{r}
property <- storm1 %>% group_by(EVTYPE) %>%  summarize(Prop.dmg=sum(PROPDMG))

names(property) <- c("Event", "Prop.damage")
property <- property[order(property$Prop.damage,decreasing=TRUE),]

crop <- storm2 %>% group_by(EVTYPE) %>%  summarize(Crop.dmg=sum(CROPDMG))

names(crop) <- c("Event", "Crop.damage")
crop <- crop[order(crop$Crop.damage,decreasing=TRUE),]

crop.prop <- left_join(crop,property,by=c("Event"= "Event"))
crop.prop$Total <- crop.prop$Crop.damage + crop.prop$Prop.damage

crop.prop <- crop.prop[order(crop.prop$Total,decreasing=TRUE),]
```


#Results

## Fatalities and injuries

### We combined the fatalities and injuries caused by weather events. Given that injuries are less impactful compared to fatalities, injuries were given a factor of 0.5, i.e. two injuries have the same impact on human health as one fatality.

```{r}
top.health <- fat.inj %>% slice(1:10)
ggplot(top.health, aes(x = reorder(Event, Combined), y = Combined)) + 
    geom_bar(stat = "identity",fill="red") + coord_flip() + 
     labs(x = "Event types", y = "Combined", title = "Top 10 events with highest impact on human health" )
```

### Tornado is the most impactful event to human health, with the excessive heat coming in far behind at second place.

## Most damaging events to property and crops

### We sum up the damage to both property and crop and identified the top 10 most damaging events.
```{r}
top.dmg <- crop.prop %>% slice(1:10)
ggplot(top.dmg, aes(x = reorder(Event, Total), y = Total)) + 
    geom_bar(stat = "identity",fill="red") + coord_flip() + 
     labs(x = "Event types", y = "Total", title = "Top 10 events with highest property & crop damage" )
```

### Floods are the most damaging weather event, almost doubling the damage done by hurricanes/typhoons in second place.